Solar-Electric and Gas Powered, Long-Endurance UAV Sizing via Geometric Programming

### 2017-05-08 docs/paperin.tex
final edits

$L$ \> atmospheric lapse rate \\ % [K/m] \\

@@

$p$ \> air pressure \\ % [Pa]\\
$p_{11}$ \> air pressure at 11,000 meters \\ % [Pa]\\

@@

$Re_{\text{h}}$ \> horizontal tail Reynolds number \\
$Re_{\text{v}}$ \> vertical tail Reynolds number \\

@@

$T_{11}$ \> air temperature at 11,000 meters \\ % [K] \\
$T_{\text{atm}}$ \> air temperature at flight altitude \\ % [K] \\
$T_{\text{sl}}$ \> air temperature at sea level \\ % [K] \\

@@

$W_{\text{empennage}}$ \> empennage weight \\ % [N] \\

@@

$\mu$ \> air viscosity \\ % [kg/m/s] \\
$\mu_0$ \> reference air viscosity \\ % [kg/m/s] \\

@@

A comparision of the gas and solar-elecrtric powered aircraft is achieved by comparing their respective capabilities at satisfying a common set of requirements shown in Table~\ref{t:mreqs}.  These requirements can and will be changed in the results section to observe which archicture is best suited to meet various sets of requirements. A discussion of each requirement is included in this section.

\begin{longtable}{lccccccccccccc}
\caption{Mission Requirements}\\
\toprule
\toprule
\label{t:mreqs}
Payload & 10 lbs\\
Station Keeping & 90\% \\
Endurance & $>6$ days\\
Season & all seasons\\
Altitude & $>$ 4,600 m\\
Latitude & $\pm30^{\circ}$\\
\bottomrule
\end{longtable}

@@

\subsection{Endurance}

For long-endurance gas powered aircraft, the endurance requirement is defined by amount of time the aircraft needs to remain in the air without refueling.  
For a solar-electric powered aircraft, however, if it can fly during the shortest day of the year, December 21st (June 21st for Southern Hemisphere), then it can thoretically fly at any time of the year, more than satisfying the 6 day endurance requirement. 
Therefore, to meet the endurance requirement the solar-electric powered aircraft must be able to fly during the winter solstice.  
To achieve this design constraint the aircraft must be able to fly with limited solar energy, discussed in Section IV.A, and with higher than average wind speeds as shown in Figure~\ref{f:windvsmonth}.  
Because becuase wind speeds are greatest during the winter solstice for both the Northern and Southern Hemispheres only winds speeds from December and June respectively, will be used for the aircraft sizing and performance analysis presented in this paper.

\begin{figure}[H]
 \begin{subfigmatrix}{2}% number of columns
     \subfigure[-30$^{\circ}$ latitude \label{f:windvsmonthS30}]{\includegraphics[natwidth=560,natheight=420]{windvsmonthS30.eps}}
     \subfigure[+30$^{\circ}$ latitude \label{f:windvsmonthN30}]{\includegraphics[natwidth=560,natheight=420]{windvsmonthN30.eps}}
 \end{subfigmatrix}
 \caption{\textbf{Winds peak at the winter solstice in Southern and Northern Hemispheres.\cite{wind}}}
 \label{f:windvsmonth}
\end{figure}

@@

Long-endurance aircraft tend to fly at low velocities, which based on Figure~\ref{f:altvswindhmin} means there are two possible operating altitude regiems: at or around the minimum payload requirement 4,600 - 7,600 m or at high altitudes between 16,700-20,000 m. 
Gas powered aircraft generally fly in the lower altitude regime of 4,600-7,600 m because naturally aspirated engines lose power with increased altitude and are unable to reach altitudes higher than about 13,500 m.  
Solar-electric powered aircraft, which do not have naturally aspirated engines, will to fly around 18,000 m to avoid cloud coverage.

\begin{figure}[H]
	\begin{center}
	\includegraphics[width=0.6\textwidth,natwidth=538,natheight=405]{altvswindhmin.eps}
    \caption{\textbf{Aircraft must fly above $h_{\text{min}}$ to meet altitude requirement. Bands represent 80th, 90th and 95th percentiles.\cite{wind}}}
	\label{f:altvswindhmin}
	\end{center}
\end{figure}

@@

It is assumed that long-endurance aircraft will have a requirement to fly within a band of latitudes (i.e. $\pm$45 degrees latitude).  
For example, an aircraft designed or optimized for the 35th latitude would be able to operate at any latitude between $\pm35$ degrees latitude. 
Latitude affects both the solar-electric and the gas powered aircraft because wind speed varies with latitude. 
Figure~\ref{f:latvswind} shows how the wind speed varies with latitude in the Northern Hemisphere in December.

\begin{figure}[H]
	\begin{center}
	\includegraphics[width=0.6\textwidth,natwidth=576,natheight=432]{latvswindN.eps}
    \caption{\textbf{Wind speeds by latitude.  Bands represent 80th, 90th and 95th percentile winds.\cite{wind} }}
	\label{f:latvswind}
	\end{center}
\end{figure}

Latitude additionally affects the solar-electric powered aircraft because at higher latitudes there is less daylight and therefore less solar energy during the winter months. This is further discussed in Section IV.A.

@@

To evaluate the gas and solar-electric powered design space, the requirements, architecture type, physical models, and assummed values are expressed as constraints on the GP.\cite{hoburgthesis} 
Essentially, the gas and solar-electric powered geometric programming optimization models are long lists of constraints in a GP-form that can be rapidly solved using interior point methods.  
Figure~\ref{f:flowchart} shows how the optimization problem is constructed and solved. 
The following sections explain in detail the configuration, physical modeling, and input value assumptions and their resulting constraint equations that were used in both the gas and solar-electric powered optimization models.

\begin{figure}[H]
	\begin{center}
	\includegraphics[width=0.8\textwidth,natwidth=568,natheight=154]{flowchart.eps}
    \caption{\textbf{Process of constructing and solving a GP optimization problem. }}
	\label{f:flowchart}
	\end{center}
\end{figure}

@@

Various structural, aerodynamic, and weight models necesistate assumptions about the vehicle configuration.  
Common characteristics between the two models include a constant tapered wing and a conventional tail with a single tail boom extending from the wing.  
The gas powered aircraft has a fuselage to hold all of the fuel required for the mission.  
The solar-electric powered aircraft holds the batteries in the wings.  
The solar cells for the solar-electric aircraft are placed along the wing and possibly on the horizontal tail.  
A simple diagram of what each vehicle might look like is shown in Figure~\ref{f:simpleaircraft} with their corresponding weight breakdowns in Figure~\ref{f:wbreak}.
The configuration is assumed to be static for this study.

\begin{figure}[H]
	\begin{center}
	\includegraphics[width=0.9\textwidth,natwidth=447,natheight=265]{simpleaircraft.eps}
    \caption{\textbf{Blue and red are specific to the solar-electric and gas archictures respectively.  Black are shared characteristics.}}
	\label{f:simpleaircraft}
	\end{center}
\end{figure}

\begin{figure}[H]
	\begin{center}
	\includegraphics[width=0.8\textwidth,natwidth=378,natheight=335]{wbreak.eps}
   \caption{\textbf{Representative weight breakdown of both aircraft optimization models.}}
	\label{f:wbreak}
	\end{center}
\end{figure}

@@

To evaluate the size and performance of both gas and solar-electric powered aircraft, underlying physics models are used.  
Environmental, aircraft performance, and structural models are used in the optimization model to capture the effects of the requirements on the size and design of the aircraft.
Equations from these models are expressed in a GP-compatible form to enable their use in the optimization. 
Understanding the models that are used in a GP is important because the solution to a GP is only as accurate as the models that are used to construct the program.  
To obtain higher fidelity results from the optimization, higher fidelity models can be implemented.

Because wind speeds also vary with latitude, several constraints are needed to optimize altitude. 
Constraints relating wind speed and air density were generated using the data fitting technique described by Hoburg et. al\cite{fitting}.
Air density was chosen as the dependent variable instead of altitude to eliminate an extra equation relating altitude to air density. 
Altitude can be post-calculated from the optimized air density. (See Appendix A)

For example, an aircraft optimized for $\pm$30 degrees latitude will have 11 different wind speed and solar flux constraints that must be satisfied, one for each degree of latitude between 20 and 30 degrees. 
Because wind speeds are low and solar flux is higher below 20 degrees latitude, no constraints for latitudes below 20 degrees are actually included in the optimization. 
The wind speed data used to generate these equations includes December wind speeds from the Northern Hemisphere and June wind speeds from the Southern Hemisphere for the years 2005-2015.
In total, 41 wind speed constraints were generated to represent $\pm$20-60 degrees each with a log space RMS error of less than 5\%. (See Appendix B) 
% The constraint on December wind speeds and air density at the 30 degree latitude is 
% 
% \begin{align}
%     \left(\frac{V_{\text{wind}}}{V_{\text{ref}}}\right)^{7.066} &\geq 18413.2\rho^{9.3}p_{\text{wind}}^{5.51} + 9658.5\rho^{8.25}p_{\text{wind}}^{23.47} \nonumber \\
%     \label{e:windfit35}
%     &+ \num{2.55d-13}\rho^{-7.72}p_{\text{wind}}^{49.65} + \num{1.11d-16}\rho^{-9.80}p_{\text{wind}}^{10.26} + 10.493.8\rho^{7.31}p_{\text{wind}}^{226.9}
% \end{align}
% 
% where $V_{\text{ref}} = 100$ [m/s] and $p_{\text{wind}}$ is the percentile wind speed. The comparison of the fitted Equation~\eqref{e:windfit35} to the actual data is presented in Figure~\ref{f:windfitl35}. 
% 
% \begin{figure}[H]
% 	\begin{center}
% 	\includegraphics[width=0.6\textwidth,natwidth=511,natheight=405]{windfitl30.pdf}
%     \caption{\textbf{Posynomial approximation to wind speed data\cite{wind} (circles) at 35$^{\circ}$ latitude. (Log-space RMS error = 0.043.)}}
% 	\label{f:windfitl35}
% 	\end{center}
% \end{figure}

For the gas powered aircraft, the altitude is not optimized.  
Gas-powered aircraft, which are naturally aspirated, will be less efficient at higher altitudes.  
Additionally, wind speeds increase nearly linearly up to 9,100 m.  
Therefore, the altitude for gas-powered aircraft is determined as the minimum altitude necessary for the payload to be effective.\cite{orion}
For this sizing study, the altitude requirement for the gas powered aircraft is $h=4,500$ m, corresponding to a 100 km diameter footprint with a 5 degree lookup angle.

\begin{figure}[H]
	\begin{center}
	\includegraphics[width=0.7\textwidth,natwidth=572,natheight=405]{lat30.eps}
    \caption{\textbf{Solar power on Dec 21st at 30$^{\circ}$ North. 24-hour operationality achieved when red area exceeds blue, magenta and green areas. }}
	\label{f:lat30}
	\end{center}
\end{figure}

where $\eta_{\text{motor}} = 0.95$. 
Thus, for a given latitude and day, and therfore total solar energy $(E/S)_{\text{sun}}$, both $(E/S)_{\text{day}}$ and $(E/S)_{\text{twilight}}$ are functions of $(P/S)_{\text{min}}$.  
The total solar energy available $(E/S)_{\text{sun}}$, is computed prior to the optimization solve and is used to generate monomial approximations for $(E/S)_{\text{day}}$ and $(E/S)_{\text{twilight}}$ as functions $(P/S)_{\text{min}}$, one for each degree of latitude between 20 and 60 degrees latitude. 
$(E/S)_{\text{sun}}$, $(E/S)_{\text{day}}$, $(E/S)_{\text{twilight}}$, and $t_{\text{night}}$ are precomputed using the formulas in Appendix C to generate the fitted equations. 
% Figure~\ref{f:energyapprox} shows sample monomial approximations at 30 degrees latitude. 
% 
% \begin{figure}[H]
%  \begin{subfigmatrix}{2}% number of columns
%      \subfigure[Monomial Approxmiation for $(E/S)_{\text{day}}$\label{f:Benergy}]{\includegraphics[natwidth=576,natheight=432]{Benergy.pdf}}
%      \subfigure[Monomial Approximation for $(E/S)_{\text{twilight}}$\label{f:Cenergy}]{\includegraphics[natwidth=575,natheight=432]{Cenergy.pdf}}
%  \end{subfigmatrix}
%  \caption{\textbf{Monomial approximations for $(E/S)_{\text{day}}$ and $(E/S)_{\text{twilight}}$. (Log-space RMS error = 0.007 and 0.0180)}}
%  \label{f:energyapprox}
% \end{figure}

For the purposes of this design study, it is assumed that the solar cells are placed on the wing and the horizontal tail but not on the vertical tail. 
The fractional solar cell area index $f_{\text{solar}}$ is one when the solar cells completely cover the main wing and greater than one if solar cells are placed on the horizontal tail,

\begin{equation}
    \label{e:electricengine}
    P_{\text{max}} = B_{PM} m_{\text{motor}}
\end{equation}

where $m_{\text{motor}}$ is the motor mass, $P_{\text{max}} \geq P_{\text{oper}}$ is the maximum operating power, and the power to mass ratio is $B_{PM} = 4140.8$ [W/kg]. \\

This version comes from assuming that $\text{BSFC}$ and the power to weight ratio, $(P_{\text{shaft}}/W)$, are constant during the considered flight segment. 
One way to obtain a constant power to weight ratio is a constant velocity and constant lift coefficient.\cite{br2}
The average weight, $W_{\text{ave}}$, is assumed to be the geometric mean, defined as

    Equations~\eqref{e:brzbre} and~\eqref{e:brtaylor} are monomial and posynomial respectively and therefore GP compatible. For long-endurance aircraft, missions can last days, causing the power to weight ratio $(P_{\text{shaft}}/W)$ to vary significantly during the course of the flight.  
    Equations~\eqref{e:brzbre},~\eqref{e:brtaylor}, and~\eqref{e:slfweight} can be discretized to account for this.

\begin{align}
    \label{e:slfweightd}
    \sqrt{W_i W_{i+1}} &= \frac{1}{2} \rho_i V_i^2 C_{L_i} S \\
    \label{e:brzbred}
    z_{bre_i} &\geq \frac{P_{\text{shaft}_i}t_i \text{BSFC} g}{\sqrt{W_i W_{i+1}}}\\
    \label{e:brtaylord}
    \frac{W_{\text{fuel}_i}}{W_{i+1}} &\geq z_{bre_i} + \frac{z_{bre_i}^2}{2} + \frac{z_{bre_i}^3}{6} + \frac{z_{bre_i}^3}{24} 
    \end{align}

\subsubsection{Elliptcal Fuselage}

where $\rho_{A_{\text{cfrp}}} = 0.0975$ [g/cm$^2$], or the area density of two plys of carbon fiber.\cite{cfply}  The surface area is also used to calculate the drag assuming a skin friction based drag model,

where $W_{\text{engine-ref}} = 10$ [lbs] and $P_{\text{ref}} = 10$ [hp].  Equation~\eqref{e:powerlaw} was found using techniques described in Hoburg et. al\cite{fitting}. Equation~\eqref{e:powerlaw} is compared to the data in Figure~\ref{f:powervsweightfit}.

\begin{figure}[H]
	\begin{center}
	\includegraphics[width=0.6\textwidth,natwidth=524,natheight=405]{powervsweightfit.eps}
    \caption{\textbf{Power law fit to University of North Dakota engine weight data\cite{gasengine}. (Log-space RMS error = 0.34)}}
	\label{f:powervsweightfit}
	\end{center}
\end{figure}

Two characteristics of gas engines affect the performance of long-endurance aircraft.  
The first is brake specific fuel consumption, $\text{BSFC}$; a lower $\text{BSFC}$ will result in increased endurance.  
The second is the lapse rate.  
Assuming a propeller driven aircraft and a naturally aspirated engine, as the aircraft reaches higher altitudes, the engine will have decreased available power. 
To account for these two affects a two-stroke, double cylinder engine, the DF70 from RCV Engines Ltd, England, was selected as a representative engine.  
RCV Engines Ltd provided manufacturing data that was used to generate representative performance curves.\cite{rcvengines}
It is assumed that engines of a similar size will perform similarly to the DF70 engine.  
This engine performance model is only valid for internal combustion engines.

To capture throttling effects as aircraft weight decreases, a GP-compatible curve was generated from the DF70 data to relate $\text{BSFC}$ and shaft power,

\begin{equation}
    \label{e:rpmtobsfc}
    \left( \frac{\text{BSFC}}{\text{BSFC}_{100\%}} \right)^{18.5563} \geq \num{8.66321d-3}\left( \frac{P_{\text{shaft}}}{P_{\text{max}}}\right)^{-7.70161} + 1.38628\left( \frac{P_{\text{shaft}}}{P_{\text{max}}} \right)^{1.12921} \\
\end{equation}

where $\text{BSFC}_{100\%} = 0.3162$ [kg/kW/hr].\cite{rcvengines}
Using this approximation the required shaft power determines the $\text{BSFC}$.
A comparison of the GP-compatible approximation to the manufacturing data from RCV engines is presented in Figure~\ref{f:powervsweightfit}. A comparison of the DF70 data provided by RCV Engine Ltd in Figure~\ref{f:powervsweightfit} to the $\text{BSFC}$ to power curves in Goering\cite{bsfcperf} verify that this performance curve is representative of engines of similar sizes.

\begin{figure}[H]
	\begin{center}
	\includegraphics[width=0.6\textwidth,natwidth=530,natheight=414]{powertobsfcfit.eps}
    \caption{\textbf{Representative engine performance fit based on RCV Engine Ltd data.  (Log-space RMS error = 0.007) }}
	\label{f:powervsweightfit}
	\end{center}
\end{figure}

\begin{equation}
    \label{e:climbrate}
    \dot{h} \geq 30 \text{ [m/min]}.
\end{equation}

\begin{equation}
    \label{e:climb}
    T \geq \frac{1}{2} C_D \rho V^2 S + W \frac{\dot{h}}{V}
\end{equation}

@@

    where the propulsive efficiency is assumed to be a constant, $\eta_{\text{prop}} = 0.75$.

@@

where the Oswald efficiency factor is assumed to be constant, $e=0.9$. 
The non-wing drag coefficient $C_{d_0}$, is estimated as a drag build up from the fuselage, tail boom, and horizontal and vertical surfaces discussed later.

    To estimate the wing profile drag, a posynomial fit is made of the drag polar of the selected \emph{JH01} airfoil (See Appendix D for a discussion of the airfoil choice). 
    Drag polars were produced using XFOIL\cite{xfoil} at various Reynolds and the data was then fit to a posynomial equation using techniques described in Hoburg et. al.\cite{fitting}
    The XFOIL drag polar data is compared to the posynomial fit in Figure~\ref{f:JH01polar}.

@@

\begin{figure}[H]
	\begin{center}
	\includegraphics[width=0.6\textwidth,natwidth=536,natheight=405]{jho1polarfit1.eps}
    \caption{\textbf{Posynomial fit (solid lines) to XFOIL data (circles).  (Log-space RMS error~=~0.00489.)}}
	\label{f:JH01polar}
	\end{center}
\end{figure}

% \subsection{Comparison to Existing Platforms}
% 
% By comparing the results of the optimization model to existing gas-powered, long-endurance platforms confidence is gained in the validity of the models.  
% Known pramaters from the Orion, Predator, Global Hawk, Hermes and Penguin (Table~\ref{t:existingparams}) aircraft are inputed to the optimization model.  
% A comparison of the results from the optimization model and the actual aircraft endurance is summarized in table~\ref{t:existingplatforms} and figure~\ref{f:vehiclecomp}. The model is then solved by maximizing endurance.
% 
% \begin{longtable}{lcccc}
%     \caption{Known Parameters of Exisiting Long Endurance Aircraft}\\
%      \toprule
%      \toprule
%      \label{t:existingparams}
%      Aircraft & MTOW [lbf] & Cruise Speed [m/s] & Altitude [ft] & Payload Weight [lbf]\\
%      \midrule
%      Orion\cite{orion}      & 11,200 & 34  & 15,000 & 1,500\\
%      Predator\cite{predator}   & 2,250  & 70  & 25,000 & 450\\
%      Hermes\cite{hermes}     & 2,600  & 70  & 30,000 & 771\\
%      Global Hawk\cite{globalhawk}& 32,250 & 160 & 50,000 & 3,000\\
%      Penguin\cite{penguin}    & 47    & 22   & 10,000 & 22\\
%      \bottomrule
%  \end{longtable}
% 
%  Additionally, the parameters in table~\ref{t:compparams} are set as constants for each comparison optimization solve. 
% 
% \begin{longtable}{ll}
%     \caption{Fixed Optimization Parameters} \\
%      \toprule
%      \toprule
%      Variable &  Value \\
%      \midrule
%      $\eta_{prop}$        & 0.65   \\
%      $A$                 & 25     \\
%      $f_{\text{structural}}$     & 0.35   \\
%      $\text{BSFC}$ [lb/hp/hr]    & 0.6    \\
%      \bottomrule
%      \label{t:compparams}
%  \end{longtable}
% 
%  \begin{figure}[H]
%  \begin{subfigmatrix}{2}% number of columns
%      \subfigure[MTOW Comparison\label{f:mtowvehiclecomp}]{\includegraphics[natwidth=610,natheight=642]{mtowvehiclecomp.pdf}}
%      \subfigure[Span Comparison\label{f:bvehiclecomp}]{\includegraphics[natwidth=610,natheight=642]{bvehiclecomp.pdf}}
%  \end{subfigmatrix}
%  \caption{ \textbf{Comparison of optimization results to existing platforms.  Each aircraft's unique parameters from table~\ref{t:existingparams} are set as fixed constants in the optimization model.} }
%  \label{f:vehiclecomp}
% \end{figure}
%  
% \begin{longtable}{lcccc}
% \caption{Optimization Results Comparison} \\
% \label{t:existingplatforms} \\
%      \toprule
%      \toprule
%      Aircraft & Actual Endurance [days] & Predicted Endurance [days] & Actual Span [ft] & Predicted Span [ft]\\
%      \midrule
%      Orion      & 5 & 4.4   & 132 & 146.8\\
%      Predator   & 1  & 1.7  & 55 & 39.0 \\
%      Hermes     & 1.5 & 1.2 & 50  & 47.27\\
%      Global Hawk& 1.5 & 1.0 & 131 & 100.6\\
%      Penguin    & 1 & 1.6   & 10.8  & 15.14\\
%      \bottomrule
%  \end{longtable}
% 
% 
% Because of limited information about these aircraft the optimization predictions are not exact. However, it does validate the proposition that the geometric programming optimzation model is useful for approximating the size and performance of long-endurance aircraft because the results of the optimization are within the same order of manitude as the existing aircarft.

@@

\begin{itemize}

\item{Standard Wing Bending Case}

The distributed load along the wing that is a combination of the lifting distribution and weight distribution along the wing
The wing loading distribution, $q(y)$, can be approximated as a scaling of the local chord,\cite{bending}

@@

where $W_{\text{cent}}$ is the sum of loads acting at the center of the aircraft and the safety load factor is an input $N_{\text{max}}=5$.  $N_{\text{max}}=1$ corresponds to steady level flight. The local chord for a constant tapered wing\cite{bending} is

\begin{equation}
    \label{e:localchord}
    \bar{c}(y) \equiv \frac{c(y)}{S/b} = \frac{2}{1+\lambda} \left( 1 + (\lambda - 1) \frac{2y}{b} \right)
\end{equation}

where the taper ratio is an input, $\lambda=0.5$.  Thus, the pre-computed distributed load

>>>>

% $L$ \> atmospheric lapse rate \\ % [K/m] \\

@@

% $p$ \> air pressure \\ % [Pa]\\
% $p_{11}$ \> air pressure at 11,000 meters \\ % [Pa]\\

@@

% $Re_{\text{h}}$ \> horizontal tail Reynolds number \\
% $Re_{\text{v}}$ \> vertical tail Reynolds number \\

@@

% $T_{11}$ \> air temperature at 11,000 meters \\ % [K] \\
% $T_{\text{atm}}$ \> air temperature at flight altitude \\ % [K] \\
% $T_{\text{sl}}$ \> air temperature at sea level \\ % [K] \\

@@

$W_{\text{emp}}$ \> empennage weight \\ % [N] \\

@@

% $\mu$ \> air viscosity \\ % [kg/m/s] \\
% $\mu_0$ \> reference air viscosity \\ % [kg/m/s] \\

@@

A comparison of the gas and solar-electric powered aircraft is achieved by comparing their respective capabilities at satisfying a common set of requirements shown in Table~\ref{t:mreqs}.  These requirements can and will be changed in the results section to observe which architecture is best suited to meet various sets of requirements. A discussion of each requirement is included in this section.

\begin{longtable}{lccccccccccccc}
\caption{Mission Requirements}\\
\toprule
\toprule
\label{t:mreqs}
Payload & 10 lbs\\
Station Keeping & 90\% winds \\
Endurance & $>6$ days\\
Season & all seasons\\
Altitude & $>$ 4,600 m\\
Latitude & $\pm30^{\circ}$\\
\bottomrule
\end{longtable}

@@

To meet the season requirement, the solar-electric aircraft must be able to fly a full day/night cycle on the winter solstice.  
If it can do so, it automatically exceeds the endurance requirement. 
To achieve this design constraint the aircraft must be able to fly with limited solar energy, discussed in Section V.B, and with higher than average wind speeds as shown in Figure~\ref{f:windvsmonth}.

\begin{figure}[h!]
 \begin{subfigmatrix}{2}% number of columns
     \subfigure[+30$^{\circ}$ latitude \label{f:windvsmonthN30}]{\includegraphics[natwidth=560,natheight=420]{windvsmonthN30.pdf}}
     \subfigure[-30$^{\circ}$ latitude \label{f:windvsmonthS30}]{\includegraphics[natwidth=560,natheight=420]{windvsmonthS30.pdf}}
 \end{subfigmatrix}
 \caption{\textbf{Winds peak at the winter solstice in Southern and Northern Hemispheres.\cite{wind}}}
 \label{f:windvsmonth}
\end{figure}

For long-endurance gas powered aircraft, the endurance requirement is the amount of time the aircraft needs to remain in the air without refueling.  
This is expected to be mostly independent of season, except insofar as the wind speeds depends on season.

Because wind speeds are greatest during the winter solstice for both the Northern and Southern Hemispheres, only wind speeds from December and June, respectively, will be used for the both the gas and solar-electric powered aircraft sizing and performance analysis.

@@

Long-endurance aircraft tend to fly at low speeds, which based on Figure~\ref{f:altvswindhmin} means there are two possible operating altitude regiems: at or around the minimum height requirement of 4,600 m, or at high altitudes between 16,700-20,000 m. 
Gas powered aircraft generally fly in the lower altitude regime of 4,600 m because naturally aspirated engines lose power with increased altitude and are unable to reach altitudes higher than about 13,500 m.  
Solar-electric powered aircraft, which do not have naturally aspirated engines, will fly around 18,000 m to avoid cloud coverage.

\begin{figure}[h!]
	\begin{center}
	\includegraphics[width=0.6\textwidth,natwidth=538,natheight=405]{altvswindhmin.pdf}
    \caption{\textbf{Aircraft must fly above $h_{\text{min}}$ to meet altitude requirement. Bands represent 80th, 90th and 95th percentiles.\cite{wind}}}
	\label{f:altvswindhmin}
	\end{center}
\end{figure}

@@

It is assumed that long-endurance aircraft will have a requirement to be capable of operating anywhere within a band of latitudes.  
For example, an aircraft designed or optimized for the 35th latitude would be able to operate at any latitude between $\pm35$ degrees. 
Latitude affects both the solar-electric and the gas powered aircraft because wind speed varies with latitude. 
Figure~\ref{f:latvswind} shows how the wind speed varies with latitude in the Northern Hemisphere in December.

\begin{figure}[h!]
	\begin{center}
	\includegraphics[width=0.6\textwidth,natwidth=576,natheight=432]{latvswindN.pdf}
    \caption{\textbf{Wind speeds by latitude.  Bands represent 80th, 90th and 95th percentile winds.\cite{wind} }}
	\label{f:latvswind}
	\end{center}
\end{figure}

Latitude additionally affects the solar-electric powered aircraft because at higher latitudes there is less daylight and therefore less solar energy during the winter months. This is further discussed in Section V.B.

@@

To evaluate the gas and solar-electric powered design space, the requirements, architecture type, physical models, and assumed parameter values are expressed as constraints on the GP.\cite{hoburgthesis} 
Essentially, the gas and solar-electric powered geometric programs are long lists of constraints in a GP-form that can be rapidly solved using interior point methods.  
Figure~\ref{f:flowchart} shows how the optimization problem is constructed and solved. 
The following sections explain in detail the configuration, physical modeling, and input value assumptions and their resulting constraint equations that were used in both the gas and solar-electric powered optimization models.

\begin{figure}[h!]
	\begin{center}
	\includegraphics[width=0.8\textwidth,natwidth=568,natheight=154]{flowchart.eps}
    \caption{\textbf{Process of constructing and solving a GP optimization problem. }}
	\label{f:flowchart}
	\end{center}
\end{figure}

@@

For this study a simple fixed-wing tractor configuration was chosen as the baseline for both architectures. 
Common characteristics between the gas and solar models include a constant tapered wing and a conventional tail with a single tail boom extending from the wing.  
The gas powered aircraft has a fuselage to hold all of the fuel required for the mission.  
The solar-electric powered aircraft holds the batteries in the wings.  
The solar cells for the solar-electric aircraft are placed along the wing and possibly on the horizontal tail.  
A simple diagram of each vehicle is shown in Figure~\ref{f:simpleaircraft} with their corresponding weight breakdowns in Figure~\ref{f:wbreak}.
The configuration is assumed to be static for this study.

\begin{figure}[h!]
	\begin{center}
	\includegraphics[width=0.9\textwidth,natwidth=447,natheight=265]{simpleaircraft.eps}
    \caption{\textbf{Blue and red are specific to the solar-electric and gas architectures respectively.  Black are shared characteristics.}}
	\label{f:simpleaircraft}
	\end{center}
\end{figure}

\begin{figure}[h!]
	\begin{center}
	\includegraphics[width=0.8\textwidth,natwidth=378,natheight=335]{wbreak.eps}
   \caption{\textbf{Representative weight breakdown of both aircraft optimization models.}}
	\label{f:wbreak}
	\end{center}
\end{figure}

@@

To evaluate the size and performance of both gas and solar-electric powered aircraft, underlying physics models are used.  
Environmental, aircraft performance, and structural models are used in the optimization model to capture the effects of the requirements on the size and design of the aircraft.
Equations from these models are expressed in a GP-compatible form to enable their use in the optimization. 
Understanding the models that are used in a GP is important because the solution to a GP is only as accurate as the models that are used to construct the program.  
To obtain higher fidelity results from the optimization, more detailed models can be implemented.

@@

    where the propulsive efficiency is assumed to be a constant, $\eta_{\text{prop}} = 0.8$.

@@

where the span efficiency factor is assumed to be constant, $e=0.9$. 
The non-wing drag coefficient $C_{d_0}$, is estimated as a drag build up from the fuselage, tail boom, and horizontal and vertical surfaces discussed later.

    To estimate the wing profile drag, a posynomial fit is made of the drag polar of the selected \emph{JH01} airfoil (See Appendix A for a discussion of the airfoil choice). 
    Drag polars were produced using XFOIL\cite{xfoil} at various Reynolds and the data was then fit to a posynomial equation. (See Appendix B)
    The XFOIL drag polar data is compared to the posynomial fit in Figure~\ref{f:JH01polar}.

@@

\begin{figure}[h!]
	\begin{center}
	\includegraphics[width=0.6\textwidth,natwidth=536,natheight=405]{jho1polarfit1.eps}
    \caption{\textbf{Posynomial fit (solid lines) to XFOIL data (circles).  (Log-space RMS error~=~0.00489.)}}
	\label{f:JH01polar}
	\end{center}
\end{figure}

% \subsection{Comparison to Existing Platforms}
% 
% By comparing the results of the optimization model to existing gas-powered, long-endurance platforms confidence is gained in the validity of the models.  
% Known parameters from the Orion, Predator, Global Hawk, Hermes and Penguin (Table~\ref{t:existingparams}) aircraft are inputed to the optimization model.  
% A comparison of the results from the optimization model and the actual aircraft endurance is summarized in table~\ref{t:existingplatforms} and figure~\ref{f:vehiclecomp}. The model is then solved by maximizing endurance.
% 
% \begin{longtable}{lcccc}
%     \caption{Known Parameters of Exisiting Long Endurance Aircraft}\\
%      \toprule
%      \toprule
%      \label{t:existingparams}
%      Aircraft & MTOW [lbf] & Cruise Speed [m/s] & Altitude [ft] & Payload Weight [lbf]\\
%      \midrule
%      Orion\cite{orion}      & 11,200 & 34  & 15,000 & 1,500\\
%      Predator\cite{predator}   & 2,250  & 70  & 25,000 & 450\\
%      Hermes\cite{hermes}     & 2,600  & 70  & 30,000 & 771\\
%      Global Hawk\cite{globalhawk}& 32,250 & 160 & 50,000 & 3,000\\
%      Penguin\cite{penguin}    & 47    & 22   & 10,000 & 22\\
%      \bottomrule
%  \end{longtable}
% 
%  Additionally, the parameters in table~\ref{t:compparams} are set as constants for each comparison optimization solve. 
% 
% \begin{longtable}{ll}
%     \caption{Fixed Optimization Parameters} \\
%      \toprule
%      \toprule
%      Variable &  Value \\
%      \midrule
%      $\eta_{prop}$        & 0.65   \\
%      $A$                 & 25     \\
%      $f_{\text{structural}}$     & 0.35   \\
%      $\text{BSFC}$ [lb/hp/hr]    & 0.6    \\
%      \bottomrule
%      \label{t:compparams}
%  \end{longtable}
% 
%  \begin{figure}[H]
%  \begin{subfigmatrix}{2}% number of columns
%      \subfigure[MTOW Comparison\label{f:mtowvehiclecomp}]{\includegraphics[natwidth=610,natheight=642]{mtowvehiclecomp.pdf}}
%      \subfigure[Span Comparison\label{f:bvehiclecomp}]{\includegraphics[natwidth=610,natheight=642]{bvehiclecomp.pdf}}
%  \end{subfigmatrix}
%  \caption{ \textbf{Comparison of optimization results to existing platforms.  Each aircraft's unique parameters from table~\ref{t:existingparams} are set as fixed constants in the optimization model.} }
%  \label{f:vehiclecomp}
% \end{figure}
%  
% \begin{longtable}{lcccc}
% \caption{Optimization Results Comparison} \\
% \label{t:existingplatforms} \\
%      \toprule
%      \toprule
%      Aircraft & Actual Endurance [days] & Predicted Endurance [days] & Actual Span [ft] & Predicted Span [ft]\\
%      \midrule
%      Orion      & 5 & 4.4   & 132 & 146.8\\
%      Predator   & 1  & 1.7  & 55 & 39.0 \\
%      Hermes     & 1.5 & 1.2 & 50  & 47.27\\
%      Global Hawk& 1.5 & 1.0 & 131 & 100.6\\
%      Penguin    & 1 & 1.6   & 10.8  & 15.14\\
%      \bottomrule
%  \end{longtable}
% 
% 
% Because of limited information about these aircraft the optimization predictions are not exact. However, it does validate the proposition that the geometric programming optimzation model is useful for approximating the size and performance of long-endurance aircraft because the results of the optimization are within the same order of manitude as the existing aircarft.

@@

%\begin{itemize}

\paragraph{\textbf{Standard Wing Bending Case}} 
The distributed load along the wing is a combination of the lifting and weight distributions along the wing.
The wing loading distribution $q(y)$, can be approximated as a scaling of the local chord,\cite{bending}

@@

where $W_{\text{cent}}$ is the sum of loads acting at the center of the aircraft and the safety load factor is an input $N_{\text{max}}=5$ ($N_{\text{max}}=1$ corresponds to steady level flight). Using the equation for the local chord of a constant tapered wing\cite{bending} with a taper ratio $\lambda=0.5$, the pre-computed distributed load

% \begin{equation}
%     \label{e:localchord}
%     \bar{c}(y) \equiv \frac{c(y)}{S/b} = \frac{2}{1+\lambda} \left( 1 + (\lambda - 1) \frac{2y}{b} \right)
% \end{equation}

% where the taper ratio is an input, $\lambda=0.5$.  Thus, the pre-computed distributed load
